Supplementary Material for Effects of bariatric surgery and weight loss on resting state functional connectivity

### 2021-02-01 manuscript/draft_rsfmri.Rmd
added abstract glassbrain plots and changes to manuscript

RYGB-induced weight loss is in part mediated by malabsorption of certain nutrients, especially fat, and reduced digestion efficiency [@Berthoud_2011] while the role of increased energy expenditure after RYGB is still unclear [@Yarmush_2017;@Lamarca_2019]. Most importantly, RYGB alters the signaling cascades between the gut, fat tissue and the brain and thereby induces profound changes in the regulation of food intake and energy homeostasis [@Berthoud_2011]. RYBG-patients frequently report altered smell and taste perception, i.e. increased sensitivity to sweet taste, show less desire to consume high-calorie and high-fat foods and actually reduce intake of such food items [@Mulla_2017, @Berthoud_2011].  These changes in appetitive and reward signaling probably go back to altered gut-brain signaling, mediated by afferent vagal nerve signaling and gut hormones like GLP-1, PYY and ghrelin. 

@@

## Quality control
We performed different quality controls for the rsfMRI data.
First, we inspected parameters of head motion (mFD) and a measure of whole-brain signal variation which is reflective of noise in the data (DVARS).
As described in @Beyer_2020, there was a group-by-time interaction on mFD. The intervention group reduced in head motion in the follow-up assessments compared to the control group. 
Further, DVARS did not qualitatively differ between groups and time points, and the correlation of DVARS and mFD remained similar over time points.
We concluded that even though there was a decrease in head motion in the intervention versus control group, the impact of head motion on imaging data remained similar over the study period in both groups.
```{r tableQcfc, echo=FALSE}
res <- read.csv("report/FDFCcorrelations.csv")
rownames(res) <- c("minimally processed", "AROMA", "AROMA + CC", "AROMA + CC + GSR")
colnames(res) <- c('mean FD-QC', 'median FD-QC', 'sig. vertex', 'sig. vertex BH', 'distance-FD-QC', 'pval')

@@

```{r calculate summary of QA, eval=FALSE, echo=FALSE}
for (i in 1:nrow(final)) {
  tmp = strsplit(toString(final[i, "scan_dir"]), '/')[[1]][8]
  conf = read.csv(
    paste0("/data/pt_02161/preprocessed/resting/detailedQA/metrics/",
      tmp, "/confounds.csv"))
  tmp = cor.test(conf$DVARS, conf$FD)
  final[i, "corr_DVARS_FD"] = tmp$estimate
  final[i, "mean_DVARS"] = mean(conf$DVARS, na.rm = TRUE)
  final[i, "max_DVARS"] = max(conf$DVARS, na.rm = TRUE)
  final
}
```

@@

```{r FDFC correlation, eval=FALSE, echo=FALSE}
# do not evaluate if saved already
source("/data/pt_02161/Analysis/Preprocessing/qa/rs_qa/group_level_QA/QC_FC_correlations/calc_qc_fc_correlations.R")
FDFC=calc_qc_fc(final)
write.csv(FDFC,"report/FDFCcorrelations.csv", row.names = FALSE)
```

```{r prepare tsnr ROI plot, eval=FALSE}
final$subj.ID_tp=paste0(final$subj.ID,'_', final$tp)

>>>>

RYGB-induced weight loss is in part mediated by malabsorption of certain nutrients, especially fat, and reduced digestion efficiency [@Berthoud_2011] while the role of increased energy expenditure after RYGB is still unclear [@Yarmush_2017;@Lamarca_2019]. Most importantly, RYGB alters the signaling cascades between the gut, fat tissue and the brain and thereby induces profound changes in the regulation of food intake and energy homeostasis [@Berthoud_2011]. RYBG-patients frequently report altered smell and taste perception, i.e. increased sensitivity to sweet taste, show less desire to consume high-calorie and high-fat foods and actually reduce intake of such food items [@Mulla_2017, @Berthoud_2011].  These changes in appetitive and reward signaling likely relate to altered gut-brain communication, mediated by afferent vagal nerve signaling and gut hormones like GLP-1, PYY and ghrelin [ @Brutman_2019]. 

@@

## Quality control
We performed different quality controls for the rsfMRI data.
First, we inspected parameters of head motion (mFD) and a measure of whole-brain signal variation which is reflective of noise in the data (DVARS).
As described in @Beyer_2020, there was a group-by-time interaction on mFD. The intervention group reduced in head motion in the follow-up assessments compared to the control group. 
Further, DVARS did not qualitatively differ between groups and time points, and the correlation of DVARS and mFD remained similar over time points.
We concluded that even though there was a decrease in head motion in the intervention versus control group, the impact of head motion on imaging data remained similar over the study period in both groups.
```{r FDFC correlation, eval=FALSE, echo=FALSE}
# do not evaluate if saved already
source("/data/pt_02161/Analysis/Preprocessing/qa/rs_qa/group_level_QA/QC_FC_correlations/calc_qc_fc_correlations.R")
FDFC=calc_qc_fc(final)
write.csv(FDFC,"report/FDFCcorrelations.csv", row.names = FALSE)
```

```{r tableQcfc, echo=FALSE}
res <- read.csv("report/FDFCcorrelations.csv")
rownames(res) <- c("minimally processed", "AROMA", "AROMA + CC", "AROMA + CC + GSR")
colnames(res) <- c('mean FD-QC', 'median FD-QC', 'sig. vertex', 'sig. vertex BH', 'distance-FD-QC', 'pval')

@@

```{r calculate summary of QA, eval=TRUE, echo=FALSE}
for (i in 1:nrow(final)) {
  tmp = strsplit(toString(final[i, "scan_dir"]), '/')[[1]][8]
  conf = read.csv(
    paste0("/data/pt_02161/preprocessed/resting/detailedQA/metrics/",
      tmp, "/confounds.csv"))
  tmp = cor.test(conf$DVARS, conf$FD)
  final[i, "corr_DVARS_FD"] = tmp$estimate
  final[i, "mean_DVARS"] = mean(conf$DVARS, na.rm = TRUE)
  final[i, "max_DVARS"] = max(conf$DVARS, na.rm = TRUE)
  final
}
write.csv(file = "report/all_QA_measures.csv", final)
```

@@

```{r prepare tsnr ROI plot, eval=TRUE}
final$subj.ID_tp=paste0(final$subj.ID,'_', final$tp)

### 2021-02-01 manuscript/supplements_rsfmri.Rmd
added abstract glassbrain plots and changes to manuscript

source('/data/gh_gr_agingandobesity_share/literature/methods/statistics/linear_models_course_rogermundry_2018/functions/glmm_stability.r')
source('/data/gh_gr_agingandobesity_share/literature/methods/statistics/linear_models_course_rogermundry_2018/functions/diagnostic_fcns.r')

```

```{r "knit setup", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex") 
options(knitr.kable.NA = '') # hide NA in tables
# set working directory of knit (RMarkdown) to directory of Rproj
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
setwd(knitr::opts_knit$get("root.dir"))
```

## Interaction effect group by time on aggregated FC (preregistered analysis)
```{r load agg FC data and prepare, echo=FALSE}
agg_FC_DMN=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/DMN/agg_FC_DMN.txt")
agg_FC_DMN_ID=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/DMN/agg_FC_DMN_ID.txt")
agg_FC_Rew_ID=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/Rew/agg_FC_Rew_ID.txt")
agg_FC_Rew=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/Rew/agg_FC_Rew.txt")
for (i in c(1:nrow(agg_FC_DMN_ID))){
  agg_FC_Rew[i,"subj.ID_tp"]=strsplit(base::strsplit(as.character(agg_FC_Rew_ID[i,"V1"]),'/')[[1]][10],'_nacc')[[1]][1]
  agg_FC_DMN[i,"subj.ID_tp"]=strsplit(base::strsplit(as.character(agg_FC_DMN_ID[i,"V1"]),'/')[[1]][10],'_pcc')[[1]][1]
}

colnames(agg_FC_Rew)=c("mean_Rew_conn","sd_Rew_conn","subj.ID_tp")
colnames(agg_FC_DMN)=c("mean_DMN_conn","sd_DMN_conn","subj.ID_tp")

final_FC=merge(final, agg_FC_Rew, by="subj.ID_tp", all.x=TRUE)
final_FC=merge(final_FC, agg_FC_DMN, by="subj.ID_tp", all.x=TRUE)
final_FC$subj.ID=as.factor(final_FC$subj.ID)
final_FC$group_factor=relevel(final_FC$group_factor, ref = "KG")

>>>>

source('/data/gh_gr_agingandobesity_share/literature/methods/statistics/linear_models_course_rogermundry_2018/functions/glmm_stability.r')
source('/data/gh_gr_agingandobesity_share/literature/methods/statistics/linear_models_course_rogermundry_2018/functions/diagnostic_fcns.r')
```

```{r "knit setup", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex")
options(knitr.kable.NA = '') # hide NA in tables
# set working directory of knit (RMarkdown) to directory of Rproj
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
setwd(knitr::opts_knit$get("root.dir"))
source('report/figures.R')
source("report/tables.R")
source("report/create_sample_df.R")
```

## Interaction effect group by time on aggregated FC (preregistered analysis)
```{r load agg FC data and prepare, echo=FALSE}
final=readRDS("/data/pt_02161/Analysis/Project2_resting_state/seed-based/Second_level /code_and_manuscript/report/tab/final.rds")

agg_FC_DMN=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/DMN/agg_FC_DMN.txt")
agg_FC_DMN_ID=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/DMN/agg_FC_DMN_ID.txt")
agg_FC_Rew_ID=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/Rew/agg_FC_Rew_ID.txt")
agg_FC_Rew=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/Rew/agg_FC_Rew.txt")
for (i in c(1:nrow(agg_FC_DMN_ID))){
  tmp=strsplit(base::strsplit(as.character(agg_FC_Rew_ID[i,"V1"]),'/')[[1]][10],'_nacc')[[1]][1]
  agg_FC_Rew[i,"subj.ID"]=strsplit(tmp,"_")[[1]][1]
  agg_FC_Rew[i,"tp"]=strsplit(tmp,"_")[[1]][2]
  tmp=strsplit(base::strsplit(as.character(agg_FC_DMN_ID[i,"V1"]),'/')[[1]][10],'_pcc')[[1]][1]
  agg_FC_DMN[i,"subj.ID"]=strsplit(tmp,"_")[[1]][1]
  agg_FC_DMN[i,"tp"]=strsplit(tmp,"_")[[1]][2]
}

colnames(agg_FC_Rew)=c("mean_Rew_conn","sd_Rew_conn","subj.ID", "tp")
colnames(agg_FC_DMN)=c("mean_DMN_conn","sd_DMN_conn","subj.ID","tp")

final_FC=merge(final, agg_FC_Rew, by=c("subj.ID","tp"), all.x=TRUE)
final_FC=merge(final_FC, agg_FC_DMN, by=c("subj.ID","tp"), all.x=TRUE)
final_FC$subj.ID=as.factor(final_FC$subj.ID)
final_FC$group_factor=relevel(final_FC$group_factor, ref = "KG")
